Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-17 typeset.tex
Typeset: remove useless double spaces in typset.
Typeset: update figure labels in typeset.

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
      from palaeo-temperature proxy records from the Greenland Ice Core
      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
      Project for Ice Coring in Antarctica \citep[EPICA,][]
      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
      anomalies from the GRIP and NGRIP records were calculated from oxygen
      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:snapshots}).

@@

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:timeseries}, upper panel), model output presents
      consistent features that can be observed across the range of forcing
      data used. In all simulations, modelled ice volumes remain relatively
      low during most of the glacial cycle, except during two major glacial
      events which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and
      between 23.2 and 16.8\,\unit{ka} during MIS~2
      (Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
      consistently reached between 53.0 and 41.3\,\unit{ka} during
      MIS~3. However, the magnitude and precise timing of these three events
      depend significantly on the choice of proxy record used to derive
      a~time-dependent climate forcing (Table~\ref{tab:extrema}).

@@

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last ice volume maximum between
      19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
      oceanic palaeo-temperature records attain their maximum ice volumes
      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
      simulation predicts an early maximum in ice volume at 23.2\,\unit{ka},
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records.  Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:timeseries}).

@@

      Despite large differences in the timing of attained volume extrema
      (Table~\ref{tab:extrema}), all model runs show relatively consistent
      patterns of glaciation. During MIS~4, all simulations produce an
      extensive ice sheet, covering an area of at least half of that
      attained during MIS~2 (Table~\ref{tab:extrema};
      Fig.~\ref{fig:snapshots}, upper panels).  Corresponding maximum ice
      volumes also differ significantly between model runs, and vary between
      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
      Table~\ref{tab:extrema}).

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:snapshots}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over
      a~wide range between 1.69 and
      2.88\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).

      Modelled ice sheet geometries during the LGM (MIS~2;
      Fig.~\ref{fig:snapshots}, lower panels) invariably include
      a~ca. 1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
      located along the spine of the Rocky Mountains. Although the
      similarity of modelled ice extents is a~direct result of the choice of
      scaling factors applied to different palaeo-temperature proxy records
      (Table~\ref{tab:records}), it is interesting to note that modelled
      maximum ice volumes also fall within a~tight range of 8.40 to
      8.91\,\unit{m}\,s.l.e.  (Table~\ref{tab:extrema}).

@@

      Independently of the palaeo-temperature records used to force the ice
      sheet model, our simulations consistently produce two glacial maxima
      during the last glacial cycle. The first maximum configuration is
      obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:timeseries},
      \ref{fig:snapshots}; Table~\ref{tab:extrema}). These events broadly
      correspond in timing to the Gladstone (MIS~4) and McConnell (MIS~2)
      glaciations documented by geological evidence for the northern sector
      of the Cordilleran ice sheet \citep{Duk-Rodkin.etal.1996,
      Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2014}, and to the
      Fraser Glaciation (MIS~2) documented for its southern sector
      \citep{Porter.Swanson.1998, Margold.etal.2014}.  There is
      stratigraphical evidence for an MIS~4 glaciation in British Columbia
      \citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014},
      but their extent and timing are still highly conjectural
      \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

      Among the simulations presented here, only those forced with the GRIP,
      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
      sheet maximum extents that may be compatible with these field
      constraints (Fig.~\ref{fig:timeseries}, lower panel;
      Table~\ref{tab:extrema}).  Simulations driven by the NGRIP, ODP~1012
      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
      maximum Cordilleran ice sheet volumes that pre-date field-based
      constraints by several thousands of years (about 6, 6 and 4\,\unit{ka}
      respectively).  Concerning the simulations driven by oceanic records,
      this early deglaciation is caused by an early warming present in the
      alkenone palaeo-temperature reconstructions
      (Fig.~\ref{fig:timeseries}, upper panel;
      \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
      a~local effect, corresponding to a~weakening of the California current
      \citep{Herbert.etal.2001}. The California current, driving cold waters
      southwards along the south-western coast of North America, has been
      shown to have weakened during each peak of global glaciation (in
      SPECMAP) during the past 550\,\unit{ka}, including the LGM, resulting
      in paradoxically warmer sea-surface temperatures at the locations of
      the ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.

      Because most of the marine margin of the Cordilleran ice sheet
      terminated in a~sector of the Pacific Ocean unaffected by variations
      in the California current, it probably remained insensitive to this
      local phenomenum. However, the above paradox illustrates the
      complexity of ice-sheet feedbacks on regional climate, and
      demonstrates that, although located in the neighbourhood of the
      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
      cannot be used as a~realistic forcing to model the Cordilleran ice
      sheet through the last glacial cycle.  Similarly, the simulation using
      the NGRIP palaeo-temperature record depicts an early onset of
      deglaciation (Fig.~\ref{fig:timeseries}) following its last glacial
      volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
      about 6\,\unit{ka} earlier than dated evidence of the LGM advance.
      There is a~fair agreement between the EPICA and Vostok
      palaeo-temperature records, resulting in only small differences
      between the simulations driven by those records. These differences are
      not subject to further analysis further; instead we focus on
      simulations forced by palaeo-temperature records from the GRIP and
      EPICA ice cores that appear to produce the most realistic
      reconstructions of regional glaciation history, yet bearing
      significant disparities in model output. To allow for a~more detailed
      comparison against the geological record, these two simulations were
      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
      Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

@@

      During maximum glaciation, both simulations position the main
      meridional ice divide over the western flank of the Rocky Mountains
      (Figs.~\ref{fig:snapshots}, lower panels and \ref{fig:mis2}). This
      result appears to contrast with palaeoglaciological reconstructions
      for central and southern British Columbia with ice divides in a~more
      westerly position, over the western margin of the Interior Plateau
      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
      Clague.Ward.2011, Margold.etal.2013a}.  These indicate that
      a~latitudinal saddle connected ice dispersal centres in the Columbia
      Mountains with the main ice divide \citep{Ryder.etal.1991,
      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
      saddle does indeed feature in our modelling results, however, in an
      inverse configuration between the main ice divide over the Columbia
      Mountains and a~secondary divide over the southern Coast Mountains
      (Fig.~\ref{fig:mis2}).

@@

      The modelled ice sheet configurations corresponding to ice volume
      maxima during MIS~4 are more sensitive to the choice of atmospheric
      forcing than those corresponding to ice volume maxima during MIS~2
      (Figs.~\ref{fig:snapshots}, upper panels and \ref{fig:mis4}). The GRIP
      simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
      maximum ice sheet extent that closely resembles that obtained during
      MIS~2, with the only major difference of being slightly less extensive
      across northern and eastern sectors. In contrast, the EPICA simulation
      produces a~lower ice volume maximum (Fig.~\ref{fig:timeseries}), which
      translates in the modelled ice sheet geometry into a~significantly
      reduced southern sector, more restricted ice cover in northern and
      eastern sectors, and generally lower ice surface elevations in the
      interior (Fig.~\ref{fig:mis4}, right panel).  Thus, only the GRIP
      simulation can explain the presence of MIS~4 glacial deposits in the
      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
      marine sediment record offshore Vancouver Island at
      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

@@

      For the Cordilleran ice sheet, geological evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
      LGM maximum extent was short-lived. To compare this finding to our
      simulations, we use numerical modelling output to compute durations of
      ice cover throughout the last glacial cycle (Fig.~\ref{fig:duration}).

      The resulting maps show that, during most of the glacial cycle,
      modelled ice cover is restricted to disjoint ice caps centred on major
      mountain ranges of the North American Cordillera
      (Fig.~\ref{fig:duration}, blue areas). A~2500\,\unit{km}-long
      continuous expanse of ice, extending from the Alaska Range in the
      northeast to the Rocky Mountains in the southwest, is only in
      operation for at most 32\,\unit{ka}, which is about a~third of the
      timespan of the last glacial cycle (Fig.~\ref{fig:duration}, hatched
      areas). However, except for its margins in the Pacific Ocean and in
      the northern foothills of the Alaska Range, the maximum extent of the
      ice sheet is attained for a~much shorter period of time of only few
      thousand years (Fig.~\ref{fig:duration}, red areas). This result
      illustrates that the maximum extents of the modelled ice sheet during
      MIS~4 and MIS~2 were both short-lived and therefore out of balance
      with contemporary climate.

      A~notable exception to the transient character of the maximum extent
      of Cordilleran ice sheet is the northern slope of the Alaska Range,
      where modelled glaciers are confined to its foothills during the
      entire simulation period (Fig.~\ref{fig:duration}, AR). This apparent
      insensitivity of modelled glacial extent to temperature fluctuations
      results from a~combination of low precipitation, high summer
      temperature and large temperature standard deviation (PDD SD) in the
      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
      glaciation to the foothills of the mountains. This result could
      potentially explain the local distribution of glacial deposits, which
      indicates that glaciers flowing on the northern slope of the Alaska
      Range have remained small throughout the Pleistocene
      \citep{Kaufman.Manley.2004}.

@@

      It is generally believed that the Cordilleran ice sheet formed by the
      coalescence of several mountain-centred ice caps
      \citep{Davis.Mathews.1944}.  In our simulations, major ice-dispersal
      centres, visible on the modelled ice cover duration maps
      (Fig.~\ref{fig:duration}), are located over the Coast Mountains (CM),
      the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
      the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
      Elias mountains, heavily glacierized at present, host an ice cap for
      the entire length of both simulations, but that ice cap does not
      appear to be a~major feed to the Cordilleran ice sheet
      (Fig.~\ref{fig:duration}, WSEM).  Although the Coast, Skeena and
      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
      glaciers for most of the last glacial cycle, providing durable
      nucleation centres for an ice sheet, this is not the case for the
      Selwyn and Mackenzie mountains (SMKM), where ice cover on the highest
      peaks is limited to a~small fraction of the last glacial cycle. In
      other words, the Selwyn and Mackenzie mountains only appear as
      a~secondary ice-dispersal centre during the coldest periods of the
      last glacial cycle. The Northern Rocky Mountains
      (Fig.~\ref{fig:duration}, NRM) do not act as a~nucleation centre, but
      rather as a~pinning point for the Cordilleran ice sheet margin coming
      from the west.

      Perhaps the most striking feature displayed by the distributions of
      modelled ice cover is the persistence of the Skeena Mountains ice cap
      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
      its predominance over the other ice-dispersal centres
      (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of
      the applied forcing, this ice cap appears to survive MIS~3
      (Fig.~\ref{fig:snapshots}, middle panels), and serves as a~nucleation
      centre at the onset of the glacial readvance towards the LGM
      (MIS~2). This situation appear similar to the neighbouring Laurentide
      ice sheet, for which the importance of residual ice for the glacial
      history leading up to the LGM has been illustrated by the MIS~3
      residual ice bodies in northern and eastern Canada as nucleation
      centres for a~much more extensive MIS~2 configuration
      \citep{Kleman.etal.2010}.

@@

      A~correlation is observed between the modelled duration of warm based
      ice cover (Fig.~\ref{fig:warmbase}) and the degree of glacial
      modification of the landscape (mainly in terms of the development of
      deep glacial valleys and troughs). We find evidence for this on the
      slopes of the Coast Mountains, the the Columbia and Rocky mountains,
      the Wrangell and Saint Elias mountains, and radiating off the Skeena
      Mountains (Figs.~\ref{fig:duration} and \ref{fig:warmbase};
      \citealp[Fig.~2]{Kleman.etal.2010}).  The Skeena Mountains, for
      example, indeed bear a~strong glacial imprint that indicates ice
      drainage in a~system of distinct glacial troughs emanating in a~radial
      pattern from the centre of the mountain range
      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
      (Fig.~\ref{fig:duration}) associated with basal ice temperatures at
      the pressure melting point (Figs.~\ref{fig:warmbase}
      and~\ref{fig:warmfrac}) explains the large-scale glacial erosional
      imprint on the landscape. A~well-developed network of glacial valleys
      running to the north-west on the west slope of the Selwyn and
      Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010};
      \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
      predominantly warm-based ice (Fig.~\ref{fig:warmfrac}). However,
      because it is only glaciated for a~short fraction of the last glacial
      cycle in our simulations (Fig.~\ref{fig:duration}), this perhaps
      indicates that the observed landscape pattern originates from multiple
      glacial cycles and witnesses an increased relative importance of the
      Selwyn and Mackenzie mountains ice dispersal centre
      (Fig.~\ref{fig:duration}, SMKM), prior to the Late Pleistocene
      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

>>>>

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
      from palaeo-temperature proxy records from the Greenland Ice Core
      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
      Project for Ice Coring in Antarctica \citep[EPICA,][]
      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
      anomalies from the GRIP and NGRIP records were calculated from oxygen
      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

@@

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents
      consistent features that can be observed across the range of forcing
      data used. In all simulations, modelled ice volumes remain relatively
      low during most of the glacial cycle, except during two major glacial
      events which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and
      between 23.2 and 16.8\,\unit{ka} during MIS~2
      (Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
      consistently reached between 53.0 and 41.3\,\unit{ka} during
      MIS~3. However, the magnitude and precise timing of these three events
      depend significantly on the choice of proxy record used to derive
      a~time-dependent climate forcing (Table~\ref{tab:extrema}).

@@

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last ice volume maximum between
      19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
      oceanic palaeo-temperature records attain their maximum ice volumes
      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
      simulation predicts an early maximum in ice volume at 23.2\,\unit{ka},
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records. Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:lr_ts}).

@@

      Despite large differences in the timing of attained volume extrema
      (Table~\ref{tab:extrema}), all model runs show relatively consistent
      patterns of glaciation. During MIS~4, all simulations produce an
      extensive ice sheet, covering an area of at least half of that
      attained during MIS~2 (Table~\ref{tab:extrema};
      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
      volumes also differ significantly between model runs, and vary between
      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
      Table~\ref{tab:extrema}).

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over
      a~wide range between 1.69 and
      2.88\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).

      Modelled ice sheet geometries during the LGM (MIS~2;
      Fig.~\ref{fig:lr_maps}, lower panels) invariably include
      a~ca. 1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
      located along the spine of the Rocky Mountains. Although the
      similarity of modelled ice extents is a~direct result of the choice of
      scaling factors applied to different palaeo-temperature proxy records
      (Table~\ref{tab:records}), it is interesting to note that modelled
      maximum ice volumes also fall within a~tight range of 8.40 to
      8.91\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).

@@

      Independently of the palaeo-temperature records used to force the ice
      sheet model, our simulations consistently produce two glacial maxima
      during the last glacial cycle. The first maximum configuration is
      obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
      \ref{fig:lr_maps}; Table~\ref{tab:extrema}). These events broadly
      correspond in timing to the Gladstone (MIS~4) and McConnell (MIS~2)
      glaciations documented by geological evidence for the northern sector
      of the Cordilleran ice sheet \citep{Duk-Rodkin.etal.1996,
      Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2014}, and to the
      Fraser Glaciation (MIS~2) documented for its southern sector
      \citep{Porter.Swanson.1998, Margold.etal.2014}. There is
      stratigraphical evidence for an MIS~4 glaciation in British Columbia
      \citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014},
      but their extent and timing are still highly conjectural
      \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

      Among the simulations presented here, only those forced with the GRIP,
      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
      sheet maximum extents that may be compatible with these field
      constraints (Fig.~\ref{fig:lr_ts}, lower panel;
      Table~\ref{tab:extrema}). Simulations driven by the NGRIP, ODP~1012
      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
      maximum Cordilleran ice sheet volumes that pre-date field-based
      constraints by several thousands of years (about 6, 6 and 4\,\unit{ka}
      respectively). Concerning the simulations driven by oceanic records,
      this early deglaciation is caused by an early warming present in the
      alkenone palaeo-temperature reconstructions
      (Fig.~\ref{fig:lr_ts}, upper panel;
      \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
      a~local effect, corresponding to a~weakening of the California current
      \citep{Herbert.etal.2001}. The California current, driving cold waters
      southwards along the south-western coast of North America, has been
      shown to have weakened during each peak of global glaciation (in
      SPECMAP) during the past 550\,\unit{ka}, including the LGM, resulting
      in paradoxically warmer sea-surface temperatures at the locations of
      the ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.

      Because most of the marine margin of the Cordilleran ice sheet
      terminated in a~sector of the Pacific Ocean unaffected by variations
      in the California current, it probably remained insensitive to this
      local phenomenum. However, the above paradox illustrates the
      complexity of ice-sheet feedbacks on regional climate, and
      demonstrates that, although located in the neighbourhood of the
      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
      cannot be used as a~realistic forcing to model the Cordilleran ice
      sheet through the last glacial cycle. Similarly, the simulation using
      the NGRIP palaeo-temperature record depicts an early onset of
      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial
      volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
      about 6\,\unit{ka} earlier than dated evidence of the LGM advance.
      There is a~fair agreement between the EPICA and Vostok
      palaeo-temperature records, resulting in only small differences
      between the simulations driven by those records. These differences are
      not subject to further analysis further; instead we focus on
      simulations forced by palaeo-temperature records from the GRIP and
      EPICA ice cores that appear to produce the most realistic
      reconstructions of regional glaciation history, yet bearing
      significant disparities in model output. To allow for a~more detailed
      comparison against the geological record, these two simulations were
      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
      Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).

@@

      During maximum glaciation, both simulations position the main
      meridional ice divide over the western flank of the Rocky Mountains
      (Figs.~\ref{fig:lr_maps}, lower panels and \ref{fig:mis2}). This
      result appears to contrast with palaeoglaciological reconstructions
      for central and southern British Columbia with ice divides in a~more
      westerly position, over the western margin of the Interior Plateau
      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
      Clague.Ward.2011, Margold.etal.2013a}. These indicate that
      a~latitudinal saddle connected ice dispersal centres in the Columbia
      Mountains with the main ice divide \citep{Ryder.etal.1991,
      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
      saddle does indeed feature in our modelling results, however, in an
      inverse configuration between the main ice divide over the Columbia
      Mountains and a~secondary divide over the southern Coast Mountains
      (Fig.~\ref{fig:mis2}).

@@

      The modelled ice sheet configurations corresponding to ice volume
      maxima during MIS~4 are more sensitive to the choice of atmospheric
      forcing than those corresponding to ice volume maxima during MIS~2
      (Figs.~\ref{fig:lr_maps}, upper panels and \ref{fig:mis4}). The GRIP
      simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
      maximum ice sheet extent that closely resembles that obtained during
      MIS~2, with the only major difference of being slightly less extensive
      across northern and eastern sectors. In contrast, the EPICA simulation
      produces a~lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
      translates in the modelled ice sheet geometry into a~significantly
      reduced southern sector, more restricted ice cover in northern and
      eastern sectors, and generally lower ice surface elevations in the
      interior (Fig.~\ref{fig:mis4}, right panel). Thus, only the GRIP
      simulation can explain the presence of MIS~4 glacial deposits in the
      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
      marine sediment record offshore Vancouver Island at
      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

@@

      For the Cordilleran ice sheet, geological evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
      LGM maximum extent was short-lived. To compare this finding to our
      simulations, we use numerical modelling output to compute durations of
      ice cover throughout the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}).

      The resulting maps show that, during most of the glacial cycle,
      modelled ice cover is restricted to disjoint ice caps centred on major
      mountain ranges of the North American Cordillera
      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,\unit{km}-long
      continuous expanse of ice, extending from the Alaska Range in the
      northeast to the Rocky Mountains in the southwest, is only in
      operation for at most 32\,\unit{ka}, which is about a~third of the
      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
      areas). However, except for its margins in the Pacific Ocean and in
      the northern foothills of the Alaska Range, the maximum extent of the
      ice sheet is attained for a~much shorter period of time of only few
      thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This result
      illustrates that the maximum extents of the modelled ice sheet during
      MIS~4 and MIS~2 were both short-lived and therefore out of balance
      with contemporary climate.

      A~notable exception to the transient character of the maximum extent
      of Cordilleran ice sheet is the northern slope of the Alaska Range,
      where modelled glaciers are confined to its foothills during the
      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This apparent
      insensitivity of modelled glacial extent to temperature fluctuations
      results from a~combination of low precipitation, high summer
      temperature and large temperature standard deviation (PDD SD) in the
      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
      glaciation to the foothills of the mountains. This result could
      potentially explain the local distribution of glacial deposits, which
      indicates that glaciers flowing on the northern slope of the Alaska
      Range have remained small throughout the Pleistocene
      \citep{Kaufman.Manley.2004}.

@@

      It is generally believed that the Cordilleran ice sheet formed by the
      coalescence of several mountain-centred ice caps
      \citep{Davis.Mathews.1944}. In our simulations, major ice-dispersal
      centres, visible on the modelled ice cover duration maps
      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains (CM),
      the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
      the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
      Elias mountains, heavily glacierized at present, host an ice cap for
      the entire length of both simulations, but that ice cap does not
      appear to be a~major feed to the Cordilleran ice sheet
      (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
      glaciers for most of the last glacial cycle, providing durable
      nucleation centres for an ice sheet, this is not the case for the
      Selwyn and Mackenzie mountains (SMKM), where ice cover on the highest
      peaks is limited to a~small fraction of the last glacial cycle. In
      other words, the Selwyn and Mackenzie mountains only appear as
      a~secondary ice-dispersal centre during the coldest periods of the
      last glacial cycle. The Northern Rocky Mountains
      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation centre, but
      rather as a~pinning point for the Cordilleran ice sheet margin coming
      from the west.

      Perhaps the most striking feature displayed by the distributions of
      modelled ice cover is the persistence of the Skeena Mountains ice cap
      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
      its predominance over the other ice-dispersal centres
      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of
      the applied forcing, this ice cap appears to survive MIS~3
      (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
      centre at the onset of the glacial readvance towards the LGM
      (MIS~2). This situation appear similar to the neighbouring Laurentide
      ice sheet, for which the importance of residual ice for the glacial
      history leading up to the LGM has been illustrated by the MIS~3
      residual ice bodies in northern and eastern Canada as nucleation
      centres for a~much more extensive MIS~2 configuration
      \citep{Kleman.etal.2010}.

@@

      A~correlation is observed between the modelled duration of warm based
      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
      modification of the landscape (mainly in terms of the development of
      deep glacial valleys and troughs). We find evidence for this on the
      slopes of the Coast Mountains, the the Columbia and Rocky mountains,
      the Wrangell and Saint Elias mountains, and radiating off the Skeena
      Mountains (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for
      example, indeed bear a~strong glacial imprint that indicates ice
      drainage in a~system of distinct glacial troughs emanating in a~radial
      pattern from the centre of the mountain range
      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice temperatures at
      the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
      and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional
      imprint on the landscape. A~well-developed network of glacial valleys
      running to the north-west on the west slope of the Selwyn and
      Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010};
      \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
      predominantly warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However,
      because it is only glaciated for a~short fraction of the last glacial
      cycle in our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
      indicates that the observed landscape pattern originates from multiple
      glacial cycles and witnesses an increased relative importance of the
      Selwyn and Mackenzie mountains ice dispersal centre
      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

      The modelled distribution of warm-based ice cover
      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is inevitably
      affected by our assumption of a~constant, 70\,\unit{mW\,m^{-2}}
      geothermal heat flux at 3\,\unit{km} depth
      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
      west of the Mackenzie Mountains experience higher-than-average
      geothermal heat flux with measured values of ca.~80 and
      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
      therefore expect even longer durations of warm-based ice cover for
      these areas if we were to include spatially variable geothermal
      forcing in our Cordilleran ice sheet simulations.
